Copulas and their potential for ecology

### 2019-06-12 Paper.Rmd
still editing

All branches of ecology study relationships among environmental and biological variables. However, 
ubiquitously used approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but not in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is exceedingly common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more 
strongly related in their left compared to right tails, or vice versa. We describe mechanisms by which 
observed copula structure and tail dependence can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. Copula structure and tail 
dependence can also reveal causal relationships between variables. We describe consequences of copula structure, including impacts 
on extinction risk and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. (C, D) Contours 
of probability density functions (pdf) of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see section on methods for the observational hypothesis), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
pdf of a normal copula (methods for the observational hypothesis) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band 
can be computed
for any band (methods for the observational hypothesis) to describe how dependence
varies from one part of the two distributions to another. Diagonal lines
show two bands, the data in the left/lower band showing stronger
dependence than the right/upper band.
Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}

@@

A main benefit of a copula approach is 
that it can sensitively detect tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, C);
and \emph{vice versa} if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, D). Datasets of the same 
correlation can have a range of tail dependence 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for the observational hypothesis).

The goal of this paper is to explore the potential for 
applications of copulas in ecology. 
Our results provide support for the idea that environmental, ecological and evolutionary processes 
generate and transmit nontrivial copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding, for inference, and for 
applications across ecology. The work we present here has begun to test this
hypothesis, and we hope our results stimulate further
evaluation of the hypothesis by the research community.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Summary and guide for analyses presented in the text. Middle boxes
correspond to ecological datasets for which Q1 was examined (in the 
second through
fourth sections of the text, see also Table \ref{tab_data_info}). Upper boxes 
correspond to potential causes of non-normal (NN) copula structure that
were examined (in the fifth and sixth sections of the text). Lower boxes 
correspond to potential consequences we examined of NN copula structure for
ecological understanding and for applications (in the seventh and eighth
sections of the text). Arrow labels (A-D for analyses pertaining 
to causes, Q2; and X-Z for analyses pertaining to consequences, Q3) 
link to locations in the text.\label{MasterFigure}}
\end{center}
\end{figure}

@@

A fourth reason we may expect copula structure to be important
in ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, captured by copulas, of the dependence between local
quantities can influence the spatial mean or sum, and how this may influence 
higher organizational levels in eoclogy and human concerns.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may show a similar dependence, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand left-tail dependence in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have copula structure distinct from that of a Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
dependence distinct from that of a normal distribution, and in particular
do they show asymmetric tail dependence? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of normal distribution. Such 
approaches do not utilize the power of copulas, so Q1 asks whether ecological
data contain dependency information distinct from what can be modelled with
standard approaches. (Q2) If the answer to
Q1 is "yes," what are some possible causes/mechanisms of non-normal
copula structure? And (Q3) what are the consequences of non-normal copula
structure and tail dependence for our understanding of ecology and for 
applications? To address Q1 (second through fourth sections of the paper), 
we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
The fifth and sixth sections of the paper 
address Q2 using simple models. The seventh and eigths sections address
Q3 using both data and models. The ninth section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, non-normal copulas have never been used in ecology. 
We argue that benefits to ecology of using a complete 
copula description of dependence will be substantial.

@@

Datasets included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data (Table \ref{tab_data_info}). Copula structure describing 
the relationship between atmospheric weather variables such as rainfall or wind speed 
as measured in multiple locations has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [RaCA; @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for sampling locations).
Species-trait data were average species basal metabolic rate (BMR, KJ per hour) and 
body mass (grams) for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years. 
Our second population dataset comprised average annual plankton abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. 
Community-level data, obtained from the Cedar Creek Ecosystem Science Reserve, were 
plant aboveground biomass [@ccdataple120] and Shannon's diversity index (computed from plant 
species percent cover data [@ccdatapce120]) for `r nrow(d_cc[[5]])` plots [@ccdataplotse120], 
each 9m by 9m, as sampled in the years 1996-2000 and 2007, each year 
analyzed separately [@Tilman2001; @Tilman2006]. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. See Appendix \ref{SM-Data_details} for additional data details.

The environmental, species-trait, and community datasets are henceforth 
called the "bivariate" datasets in the sense that they comprise two quantities measured at 
different locations or for different species (Table \ref{tab_data_info}). 
The phenology, population, and ecosystem functioning datasets are 
"multivariate" in that they comprise, for each taxon, measurements through time at multiple locations; 
copula structure can be studied for each pair of locations.

@@

# Background, concepts and methods for testing the observational hypothesis\label{Methods}

@@

We introduce the copula families we used, and 
measures of tail dependence.
The normal copula family (henceforth N) is a one-parameter
family, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag1}G for an example).
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than an N
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}C-D) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}C) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}D)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several families of copulas exhibiting various 
lower- and upper-tail dependence characteristics 
(Clayton, survival-Clayton, Gumbel, survival-Gumbel, Joe, survival-Joe, Frank,
BB1, survival-BB1,
BB6, survival-BB6, BB7, survival-BB7, BB8, survival-BB8, henceforth
abbreviated C, SC, G, SG, J, SJ, F, BB1, SBB1, BB6, SBB6, BB7, SBB7, BB8, SBB8).
Copulas are implemented in the R packages `copula` and `VineCopula`. 
Visualizations are in 
Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8}. 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence.
See @nelsen2006_copula, @joe2014_dependence for further background.

We tested the observational hypothesis (Fig. \ref{MasterFigure}, O) 
principally via a model selection procedure in which the 
copulas listed above were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, we produced normalized ranks 
$u_i$ and $v_i$ as in the Introduction. 
Second, we tested the independence of the $u_i$ and $v_i$ 
(using `BiCopIndTest` in the `VineCopula` package). 
Our model 
selection algorithms were ineffective if data could not be distinguished from independent data 
since many of the copula families we considered include the independent copula. 
If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the 16 bivariate copula families listed previously to the 
normalized ranks via maximum likelihood, using `BiCopEst` in `VineCopula`, 
then obtaining AIC and BIC values and accompanying model weights $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. `BiCopEst` also provided the lower- and 
upper-tail dependence of the best-fitting member of each family. 
Fourth, $\text{AIC}_{\text{w}}$ values were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
and as our final step, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure (`BiCopGofTest` of `VineCopula`). The procedure performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
a run using $100$ bootstraps was performed; if
the $p$-value from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

The first part of the observational hypothesis was tested for a bivariate dataset 
by comparing the AIC weights of the non-N copula
families with that of the N family. 
The second part was tested by comparing 
the model-averaged lower- and upper-tail dependence
values to the lower- and upper-tail dependence values of an N copula (i.e., 0). 
The third part 
was tested by comparing the difference of the model-averaged lower- and upper-tail 
dependence values to the analogous difference for an N copula, again 0.
For multivariate datasets, we performed the bivariate analysis
for all possible pairwise combinations of locations, and 
compared AIC weights and tail-dependence of 
minimum-AIC copulas with that of an N copula. We also counted the number of location pairs 
for which model-averaged lower-tail dependence was 
greater than model-averaged upper-tail dependence 
or vice-versa, and reported the percentage of non-independent 
location pairs showing stronger lower-tail dependence.

In addition to our model selection approach, we also used nonparametric approaches to 
provide greater 
confidence in our tests of the observational hypothesis. We used three 
statistics which quantify the extent to which 
the normalized ranks $u_i$ and $v_i$ are 
related in any part of their distributions; we here describe the 
statistics conceptually, with details in Appendix \ref{SM-nonparam_stats}.
The statistics are defined with positively associated
variables in mind. All our variables were positively associated
when they were significantly associated.
Given two bounds 
$0 \leq l_b < u_b \leq 1$, define the lines $u+v=2l_b$ and 
$u+v=2u_b$, which intersect the unit square (Figs \ref{fig_cop_pedag1}H, 
\ref{SM-fig_stat_image}). 
Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and $u_b\leq 0.5$ gives information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ gives information about dependence in 
the right parts. The first statistic, 
$\cor_{l_b,u_b}(u,v)$, is the portion of the Spearman 
correlation of $u_i$ and $v_i$ that is attributable to the points 
between the bounds.
Larger values indicate stronger positive association.
We also defined a statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b},u_{b}}(u,v)$.
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong dependence
between the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.

@@

# Results of testing the observational hypothesis: Ecological datasets often have non-normal copula structure and asymmetric tail dependence \label{Results}

@@

Across essentially all our datasets, copula structures were typically non-N and 
commonly showed asymmetric tail 
dependence: the observational hypothesis (Fig. \ref{MasterFigure}) was supported. To make results easier to absorb, we first 
present results for an example bivariate dataset, then we summarize results
for all bivariate datasets, then we present results for an example multivariate dataset, then we summarize for all multivariate datasets.

>>>>

All branches of ecology study relationships among environmental and biological variables. However, 
ubiquitously used approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but not in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is exceedingly common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail dependence, whereby two variables are more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and tail dependence can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. Copula structure and tail 
dependence can also reveal causal relationships between variables. We describe consequences of copula structure, including impacts 
on extinction risk and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to elevate copula reasoning to widespread use in ecology.

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=3.5in]{./Results/PedagogFig1.pdf}
\caption{(A-D) Bivariate datasets showing diverse relationships between 
variables; many of these datasets nevertheless have the same Pearson (P)
or Spearman (S) correlation coefficients. (C, D) Contours 
of probability density functions (pdf) of Clayton (C) and survival-Clayton 
(D) copulas with parameter 5 (see section on methods for Q1), together
with samples from the corresponding distributions. (E,F) Reflections
of C, D, respectively, about a vertical axis. (G) Contours of the
pdf of a normal copula (methods for Q1) with 
parameter chosen to
have the same Spearman correlation as A-D, together with a sample.
(H) The portion of the Spearman correlation within a band 
can be computed
for any band (methods for Q1) to describe how dependence
varies from one part of the two distributions to another. Diagonal lines
show two bands, the data in the left/lower band showing stronger
dependence than the right/upper band.
Subscripts h and v 
stand for "horizontal" and "vertical". \label{fig_cop_pedag1}}
\end{center}
\end{figure}

@@

A main benefit of a copula approach is 
that it can sensitively detect tail dependence and asymmetries of tail dependence. 
Tail dependence is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail dependence than right- or upper-tail dependence
(Fig. \ref{fig_cop_pedag1}A, C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}B, D). Datasets of the same 
correlation can have a range of tail dependence 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see section on
methods for Q1).

The goal of this paper is to explore the potential for 
applications of copulas in ecology. 
Our results provide support for the hypothesis that environmental, ecological and evolutionary processes 
generate and transmit nontrivial copula structure and tail dependence, with substantial 
consequences for theoretical and conceptual understanding, for inference, and for 
applications across ecology. The work we present here has begun to test this
hypothesis, and we hope our results stimulate further
evaluation of the hypothesis by the research community.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Summary and guide for analyses presented in the text. Middle boxes
correspond to ecological datasets for which Q1 was examined (in the 
second through
fourth sections of the text, see also Table \ref{tab_data_info}). Upper boxes 
correspond to potential causes of non-normal (non-N) copula structure that
were examined (in the fifth and sixth sections of the text). Lower boxes 
correspond to potential consequences we examined of non-N copula structure for
ecological understanding and for applications (in the seventh and eighth
sections of the text). Arrow labels (A-D for analyses pertaining 
to causes, Q2; and X-Z for analyses pertaining to consequences, Q3) 
link to locations in the text.\label{MasterFigure}}
\end{center}
\end{figure}

@@

A fourth reason we may expect a copula approach to be important
for ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, captured by copulas, of the dependence between local
quantities can influence the spatial mean or sum, and how this may influence 
higher organizational levels in ecology and human concerns.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may be subject to a similar 
effect, with
potentially substantial consequences. For instance, if populations of a 
pest species in different locations are all positively associated and also strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand left-tail dependence in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have copula structure distinct from that of a Gaussian/normal
distribution (here called non-normal copula structure)? Do they show tail 
dependence distinct from that of a normal distribution, and in particular
do they show asymmetric tail dependence? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. Such 
approaches do not utilize the power of copulas, so Q1 asks whether ecological
data contain dependency information distinct from what can be modelled with
standard approaches. (Q2) If the answer to
Q1 is "yes," what are some possible causes/mechanisms of non-normal
copula structure? And (Q3) what are the consequences of non-normal copula
structure and tail dependence for our understanding of ecology and for 
applications? To address Q1 (the second through fourth sections of the paper), 
we analyze several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
To address Q2 (the fifth and sixth sections of the paper) we use 
simple models. To address Q3 (the seventh and eigths sections) we use 
both data and models. The ninth section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000]. 
To our knowledge, non-normal copulas have not been used in ecology. 
We argue that benefits to ecology of using a complete 
copula description of dependence will be substantial.

@@

Datasets included environmental, species-trait, phenological, 
population, community, and ecosystem functioning data (Table \ref{tab_data_info}), selected to span multiple fields and levels
of organization within ecology. Copula structure describing 
the relationship between atmospheric weather variables such as rainfall or wind speed 
as measured in multiple locations has been 
examined previously in the meteorological literature [e.g., @Serinaldi2008; @li2013]. We therefore 
examined environmental variables from the soil instead, using the Rapid Carbon 
Assessment database [RaCA; @Wills2014].
<!--***DAN: Add Wills and Loecke pending reference if published in time--> 
The database comprises measurements of soil organic carbon and total soil 
nitrogen (Mg C or N per hectare of soil surface) from `r nrow(d_soilCN)` 
locations across the coterminous United States (Fig. \ref{SM-fig_SoilCNmapUSA} for sampling locations).
Species-trait data were average species basal metabolic rate (BMR, KJ per hour) and 
body mass (grams) for `r nrow(d_birdsBMR)`  species of birds [@Mcnab_2009] 
and `r nrow(d_mammalsBMR)`  species of mammals [@Mcnab_2008].
These data have been much studied, but to our knowledge the copula structure 
of the dependence between metabolic rate and body mass has not been examined. 
Species-trait data such as these reflect the coevolution of the two traits.
Phenological data were aphid first-flight dates from 10 locations (Fig. \ref{SM-fig_aphid_map}) 
across the United Kingdom (UK) for 20 aphid species (Table \ref{SM-tab_aphid_info}), for the 
35 years 1976 to 2010. These time series were computed from the Rothamsted Insect 
Survey (RIS) suction-trap dataset [@Harrington2014; @Bell2015]. The first of our two 
population-level datasets was also derived from the RIS suction-trap data, and comprised 
total counts of aphids trapped for the same locations, species, and years. 
Our second population dataset comprised average annual plankton abundance estimates 
for 14 locations (Fig. \ref{SM-fig_plankton_map})
in the North Sea and British seas for 22 taxa (Table \ref{SM-tab_plankton_info}) for the 
56 years 1958 to 2013. The locations are $2^\circ$ by $2^\circ$ grid cells. 
Community-level data, obtained from the Cedar Creek Ecosystem Science Reserve, were 
plant aboveground biomass [@ccdataple120] and Shannon's diversity index (computed from plant 
species percent cover data [@ccdatapce120]) for `r nrow(d_cc[[5]])` plots [@ccdataplotse120], 
each 9m by 9m, as sampled in the years 1996-2000 and 2007, each year 
analyzed separately [@Tilman2001; @Tilman2006]. 
Finally, ecosystem functioning data were methane (CH$_{4}$) fluxes between the soil or 
water surface and the troposphere, measured at 13 locations at daily to weekly intervals 
from September 2015 to September 2016 at the Great Miami Wetland mitigation bank in Trotwood, 
Ohio [@Holland1999; @Jarecke2016; @smyth2019using]. Each included location was measured on at least 50 dates. See Appendix \ref{SM-Data_details} for additional data details.

Our environmental, species-trait, and community datasets happen to be 
"bivariate" datasets in the sense that they comprise two quantities measured at 
different locations or for different species (Table \ref{tab_data_info}). 
Our phenology, population, and ecosystem functioning datasets are 
"multivariate" in that they comprise, for each taxon, measurements through time at multiple locations; 
copula structure can be studied for each pair of locations.

@@

# Concepts and methods for Q1 \label{Methods}

@@

We introduce the copula families and 
measures of tail dependence we used.
The normal copula family (henceforth N) is a one-parameter
family, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag1}G for an example).
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than an N
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}C-D) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}C) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}D)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several families of copulas exhibiting various 
lower- and upper-tail dependence characteristics 
(Clayton, survival-Clayton, Gumbel, survival-Gumbel, Joe, survival-Joe, Frank,
BB1, survival-BB1,
BB6, survival-BB6, BB7, survival-BB7, BB8, survival-BB8, henceforth
abbreviated C, SC, G, SG, J, SJ, F, BB1, SBB1, BB6, SBB6, BB7, SBB7, BB8, SBB8).
Copulas are implemented in the R packages `copula` and `VineCopula`. 
Visualizations of these families (Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8})
can be compared with normalized rank plots to estimate visually 
which families may effectively model a given bivariate dataset. 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence.
See @nelsen2006_copula, @joe2014_dependence for further background.

We addressed Q1 principally via a model selection procedure in which the 
copulas listed above were fitted to 
our ecological datasets and fits were compared via the Akaike and Bayesian 
Information Criteria (AIC and BIC). For bivariate datasets 
$(x_i,y_i)$ for $i=1,...,n$, model selection involved several steps. 
First, we produced normalized ranks 
$u_i$ and $v_i$ as in the Introduction. 
Second, we tested the independence of the $u_i$ and $v_i$ 
(using `BiCopIndTest` in the `VineCopula` package). 
Our model 
selection algorithms were ineffective if data could not be distinguished from independent data 
since many of the copula families we considered include the independent copula. 
If independence could be rejected ($0.05$ significance level), model 
selection proceeded. Third, we fit the 16 bivariate copula families listed previously to the 
normalized ranks via maximum likelihood, using `BiCopEst` in `VineCopula`, 
then obtaining AIC and BIC values and accompanying model weights $\text{AIC}_{\text{w}}$ and 
$\text{BIC}_{\text{w}}$ [@burnham2003_modelselection]. `BiCopEst` also provided the lower- and 
upper-tail dependence of the best-fitting member of each family. 
Fourth, $\text{AIC}_{\text{w}}$ values were used to get
model-averaged lower- and upper-tail dependence values using standard model averaging 
formulas [@burnham2003_modelselection]; likewise for BIC.
Model selection methods give the *relative* support of several models, but do not 
indicate whether any of the models are an objectively good fit. To test that, 
and as our final step, we tested the goodness of fit of our AIC-best copula family using a 
bootstrapping procedure (`BiCopGofTest` of `VineCopula`). The procedure performed 
one test based on a Cramer-von Mises statistic and another based on a 
Kolmogorov-Smirnov statistic. To keep computation times reasonable, 
a run using $100$ bootstraps was performed; if
the $p$-value from either test was less than $0.2$, tests were re-run with 
$1000$ boostraps.

The first part of Q1 was answered for a bivariate dataset 
by comparing the AIC weights of the non-N copula
families with that of the N family. 
The second part of Q1 was answered by comparing 
the model-averaged lower- and upper-tail dependence
values to the lower- and upper-tail dependence values of an N copula (i.e., 0). 
The third part of Q1
was answered by comparing the difference of the model-averaged lower- and upper-tail 
dependence values to the analogous difference for an N copula, again 0.
For multivariate datasets, we performed the bivariate analysis
for all possible pairwise combinations of locations, and 
compared AIC weights and tail-dependence of 
minimum-AIC copulas with that of an N copula. We also counted the number of location pairs 
for which model-averaged lower-tail dependence was 
greater than model-averaged upper-tail dependence 
or vice-versa, and reported the percentage of non-independent 
location pairs showing stronger lower-tail dependence.

In addition to our model selection approach, we also used 
nonparametric approaches to provide greater 
confidence in our answer to Q1. We used three 
statistics which quantify the extent to which 
the normalized ranks $u_i$ and $v_i$ are 
related in any part of their distributions; we here describe the 
statistics conceptually, with details in Appendix \ref{SM-nonparam_stats}.
The statistics are defined with positively associated
variables in mind. All our variables were positively associated
when they were significantly associated.
Given two bounds 
$0 \leq l_b < u_b \leq 1$, define the lines $u+v=2l_b$ and 
$u+v=2u_b$, which intersect the unit square (Figs \ref{fig_cop_pedag1}H, 
\ref{SM-fig_stat_image}). 
Our statistics quantify the dependence
between $u_i$ and $v_i$ in the region bounded by these lines.
Using $l_b=0$ and $u_b\leq 0.5$ gives information about 
dependence in the left parts of the distributions of $u$ and $v$, and 
using $u_b=1$ and $l_b \geq 0.5$ gives information about dependence in 
the right parts. The first statistic, 
$\cor_{l_b,u_b}(u,v)$, is the portion of the Spearman 
correlation of $u_i$ and $v_i$ that is attributable to the points 
between the bounds.
Larger values indicate stronger positive association.
We also defined a statistic $\Ps_{l_b,u_b}$ (Appendix \ref{SM-nonparam_stats}), 
which has a similar 
interpretation to $\cor_{l_{b},u_{b}}(u,v)$.
Our third statistic, $\Dsq_{l_b,u_b}(u,v)$, is the average squared 
distance between points satisfying $u_i+v_i>2l_b$ and 
$u_i+v_i<2u_b$ and the line $v=u$.
Unlike $\cor_{l_b,u_b}$ and 
$\Ps_{l_b,u_b}$, for which large values inidicate strong dependence
between the bounds, small values of 
$\Dsq_{l_b,u_b}$ indicate strong dependence.

@@

# Results for Q1: Ecological datasets often have non-normal copula structure and asymmetric tail dependence \label{Results}

@@

Across essentially all our datasets, copula structures were typically non-N and 
commonly showed asymmetric tail 
dependence, answering Q1 in the affirmative. 
To make results easier to absorb, we first 
present results for an example bivariate dataset, then we summarize results
for all bivariate datasets, then we present results for an example multivariate dataset, then we summarize for all multivariate datasets.

<!--bivariate example-->
For the soil C and N data (Data, Table \ref{tab_data_info}, 
Fig. \ref{fig_biv_multi_raw_cop}A,E), variables were 
non-independent ($p=$ `r BivMS_res_Loecke$IndepTestRes`, to within the precision available 
from `BiCopIndTest` in the `VineCopula` package); non-independence is also visually apparent.
The Kendall correlation was `r round(BivMS_res_Loecke$TauVal,2)`. We fitted our 16 copulas
to the data and computed AIC and BIC weights and corresponding
lower- and upper-tail statistics for the fitted distributions (Table \ref{SM-tab_soilCNfit}).
`r BivMS_res_Loecke$InfCritRes$copname[which.max(BivMS_res_Loecke$InfCritRes$AICw)]` was the 
minimum-AIC copula, with AIC weight `r round(max(BivMS_res_Loecke$InfCritRes$AICw),2)`, whereas the 
N copula had AIC weight essentially 0, answering the 
first part of Q1 for these data. 
Model-averaged lower- and upper-tail dependence statistics were `r round(BivMS_res_Loecke$relLTdep_AICw,3)` 
and `r round(BivMS_res_Loecke$relUTdep_AICw,3)`, respectively (AIC weights were used for averaging). 
These values are distinct from what an N copula would give, namely, 0, answering the second part 
Q2 for these data. The values also differed substantially 
from each other, answering the third part of Q1. 
The numbers reflect stronger upper- than lower-tail
dependence, and this is also visible in the extreme upper-right
and lower-left corners of the copula plot (Fig. \ref{fig_biv_multi_raw_cop}E).
